### 2021-06-15 JSS/using_drtmle_jss.Rmd
touch-ups to censoring

The `drtmle` package contains methods that compute doubly-robust confidence
intervals. The `ci` method computes confidence intervals about the estimated
marginal means in a fitted `drtmle` object, in addition to contrasts between
those means. By default the method gives confidence intervals about each mean.

@@

Alternatively the `contrast` option allows users to specify a linear
combination of marginal means, which can be used to estimate the average
treatment effect as we show below.

```{r}
ci(npreg_fit, contrast = c(1,-1))
```

The `contrast` option can also be input as a list of functions in order to
compute a confidence interval for a user-specified contrast. For example, we
might be interested in the risk ratio $\psi_0(1) / \psi_0(0)$. When
constructing Wald style confidence intervals for ratios, it is common to
compute the interval on the log scale and back-transform. That is, the
confidence interval is of the form
$$
\mbox{exp}\biggl[\mbox{log}\biggl\{ \frac{\psi_0(1)}{\psi_0(0)} \biggr\} \pm
z_{1-\alpha/2} \sigma^{\text{log}}_n \biggr] \ ,
$$
where $\sigma^{\text{log}}_n$ is the estimated standard error on the log-scale.
By the delta method, an estimate of this standard error is given by $$
\sigma^{\text{log}}_n = \nabla g(\psi_n)^T \Sigma_n \nabla g(\psi_n),
$$
where $\Sigma_n$ is the doubly-robust covariance matrix estimate output from
`drtmle` and $\nabla g(\psi)$ is the gradient of
$\mbox{log}\{\psi(1)/\psi(0)\}$. To obtain this confidence interval, we may use
the following code.

```{r}
riskRatio <- list(f = function(eff){ log(eff) },
                  f_inv = function(eff){ exp(eff) },
                  h = function(est){ est[1]/est[2] },
                  fh_grad =  function(est){ c(1/est[1],-1/est[2]) })
ci(npreg_fit, contrast = riskRatio)
```

More generally this method of inputting the `contrast` option to the `ci`
method can be used to compute confidence intervals of the form
$$
f^{-1}\bigl\{ f(h(\psi_n)) \pm z_{1-\alpha/2} \nabla f(h(\psi_n))^T \Sigma_n
\nabla f(h(\psi_n)) \bigr\} \ ,
$$
where $f$ (`contrast$f`) is the transformation of the confidence interval,
$f^{-1}$ (`contrast$f_inv`) is the back-transformation of the interval, $h$
(`contrast$h`) is the contrast of marginal means, and $\nabla f(h(\psi_n))$
(`contrast$fh_grad`) defines the gradient of the transformed contrast at the
estimated marginal means.

We note that this method of computing confidence intervals can also be used to
obtain a transformed confidence interval for a particular marginal mean. For
example, in our running example $Y$ is binary. Thus, we might wish to enforce
that our confidence interval be computed on a scale which ensures that the
confidence limits are bounded between zero and one. To that end, we can
consider an interval of the form construct an interval on the logit scale and
back-transform,
$$
\mbox{expit}\biggl[ \mbox{log}\biggl\{ \frac{\psi_n(1)}{1 - \psi_n(1)} \biggr\}
\pm z_{1-\alpha/2} \sigma^{\text{logit}}_n \biggr] \ .
$$
This confidence interval may be implemented as follows.

```{r}
logitMean <- list(f = function(eff){ qlogis(eff) },
                  f_inv = function(eff){ plogis(eff) },
                  h = function(est){ est[1] },
                  fh_grad = function(est){ c(1/(est[1] - est[1]^2), 0) })
ci(npreg_fit, contrast = logitMean)
```

@@

As with the `ci` method, the `wald_test` method allows users to test linear
combinations of marginal means by inputting a vector of ones and negative ones
in the `contrast` option. We can use this to test the null hypothesis of no
average treatment effect or to test the null hypothesis that the average
treatment effect equals 0.05 (by specifying the `null` option)

@@

The `wald_test` method also allows for user-specified contrasts to be tested
using syntax similar to the `ci` method. The only difference syntactically is
that it is not strictly necessary to specify `f_inv`, as the hypothesis test is
performed on the transformed scale. That is, we can generally test the null
hypothesis that $f(h(\psi_0))$ equals $f_0$ (the function $f$ applied to the
value passed to `null`) using the Wald statistic $$
Z_n := \frac{\{f(h(\psi_n)) - f_0\}}{\nabla f(h(\psi_n))^T \Sigma_n \nabla
f(h(\psi_n))} \ .
$$
We can use the `riskRatio` contrast defined previously to test the null
hypothesis that $\psi_0(1)/\psi_0(0) = 1$.

@@

The `drtmle` function supports missing data in `A` and `Y`. Such missing data
are common in practice and, if ignored, can bias estimates of the marginal
means of interest. The estimators previously discussed can be modified to allow
this missingness to depend on covariates. Let $\Delta_A$ and $\Delta_Y$ be
indicators that $A$ and $Y$ are observed, respectively. We can redefine the OR
to be $\bar{Q}_0(a,w) := E_0(\Delta_Y Y \mid \Delta_A A = a, \Delta_A = 1,
\Delta_Y = 1, W = w)$ and similarly redefine the PS to be \begin{align}
g_0(a \mid w) &= Pr_0(\Delta_A = 1, \Delta_A A = a, \Delta_Y = 1 \mid W = w)
\notag \\
&= Pr_0(\Delta_A = 1 \mid W) Pr_0(\Delta_A A = a \mid \Delta_A = 1, W = w)
\label{modPS} \\
&\hspace{1.5in} Pr_0(\Delta_Y = 1 \mid \Delta_A = 1, \Delta_A A = a, W = w) \ \notag.
\end{align}
We introduce missing values to `A` and `Y` in our running example.

@@

The only modification of the call to `drtmle` is in how the PS estimation is
performed. The options `glm_g` and `SL_g` are still used to fit generalized
linear models or a super learner, respectively. However, the code allows for
separate regression fits for each of the three components of the PS in (8). To
perform separate generalized linear model fits for each of the three
components, `glm_g` should be a named list with entries `"DeltaA"`, `"A"`, and
`"DeltaY"`. Respectively these should provide a regression formula for the
regression of $\Delta_A$ on $W$, $A$ on $W$ in observations with $\Delta_A =
1$, and a regression of $\Delta_Y$ on $W$ and $A$ in observations with
$\Delta_A = \Delta_Y = 1$ (if `stratify = FALSE`) or a regression of $\Delta_Y$
on $W$ to be fit in observations with $\Delta_A = \Delta_Y = 1$ and $A = a$ for
each $a$ specified in option `a_0` (if `stratify = TRUE`). If only a single
formula is passed to `glm_g`, it is used for each of the three regressions. We
illustrate a call to `drtmle` with missing data and generalized linear model
fits below.

@@

It is possible to mix generalized linear models and super learning when fitting
each of the components of the PS. In the below example we use the wrapper
function `SL.glm`, which fits a main terms logistic regression, to fit the
regression of $\Delta_A$ on $W$, use `SL.npreg` to fit the regression of $A$ on
$W$, and use a super learner with library `SL.glm`, `SL.mean`, and `SL.earth` to
fit the regression of $\Delta_Y$ on $W$ and $A$.

@@

We can also fit the PS regressions externally and pass in via `gn`, though this
process is slightly more complicated than when no missing data are present. 
The basic idea is to fit one regression for (i) the indicator of missing $A$, 
(ii) $A$ itself, and (iii) the indicator of missing the outcome. The three PS's 
are then multiplied together and an appropriately ordered list is constructed
as before. We illustrate with the following simple example.

@@

So far in our examples, we have only considered binary treatments. However,
`drtmle` is equipped to handle treatments with an arbitrary number of discrete
values. Suppose that $A$ assumes values in a set $\mathcal{A}$, the PS
regression estimation is modified to ensure that
$$
  \sum\limits_{a \in \mathcal{A}} g_n(a \mid w) = 1 \ \mbox{for all $w$} \ .
$$
If this condition is true, we say that the estimates of the PS are compatible.
Many regression methodologies that have been developed work well with binary
outcomes, but have not been extended to multi-level outcomes. To ensure that
users of `drtmle` have access to the large number of binary regression
techniques when estimating the PS, we have taken an alternative approach to
estimation of the propensity for each level of the treatment. Specifically,
rather than fitting a single multinomial-type regression, we fit a series of
binomial regressions. As a concrete example, consider the case of no missing
data and where $\mathcal{A} = \{0,1,2\}$. We can ensure compatible estimates of
the PS by generating an estimate $g_n(0 \mid w)$ of $Pr_0(A = 0 \mid W = w)$
and $\tilde{g}_n(1 \mid w)$ of $Pr_0(A = 1 \mid A > 0, W = w)$. The latter
estimate may be generated by regression $I(A = 1)$ on $W$ in observations with
$A > 0$. Because the outcome is binary, this regression may be estimated using
any technique suitable for binary outcomes. By simple rules of conditional
probability, we know that $g_0(1 \mid w) = Pr_0(A = 1 \mid A > 0, W = w)\{1 -
Pr_0(A = 0 \mid W)\}$ and thus an estimate of $g_0(1 \mid w)$ can be computed
as $g_n(1 \mid w) = \tilde{g}_n(1 \mid w) \{1 - g_n(0 \mid w)\}$. Finally, we
let $g_n(2 \mid w) = 1 - g_n(0 \mid w) - g_n(1 \mid w)$, which ensures
compatibility of the estimates for each level of covariates. This approach
generalizes to an arbitrary number of treatment levels. Below, we provide an
illustration using simulated data with a treatment that has three levels. The
true parameter values in this simulation are $\psi_0(0)=$ `r round(EY0,2)`,
$\psi_0(1)=$ `r round(EY1, 2)`, and $\psi_0(2)=$ `r round(EY2,2)`.

>>>>

The `drtmle` package contains methods that compute doubly-robust confidence
intervals. The `ci()` method computes confidence intervals about the estimated
marginal means in a fitted `drtmle` object, in addition to contrasts between
those means. By default the method gives confidence intervals about each mean.

@@

Alternatively the `contrast` option allows users to specify a linear
combination of marginal means, which can be used to estimate the average
treatment effect, as we show below.

```{r}
ci(npreg_fit, contrast = c(1, -1))
```

The `contrast` option also accepts as input a list of functions to be used to
compute a confidence interval for a user-specified contrast. For example, we
might be interested in the risk ratio $\psi_0(1) / \psi_0(0)$. When
constructing Wald-style confidence intervals for ratios, it is common to
compute the interval on the logarithmic scale and subsequently back-transform.
That is, the confidence interval takes the form
\begin{equation*}
  \exp \biggl[\log \biggl\{ \frac{\psi_0(1)}{\psi_0(0)} \biggr\} \pm
  z_{1-\alpha/2} \sigma^{\log}_n \biggr] \ ,
\end{equation*}
where $\sigma^{\log}_n$ is the estimated standard error on the log-scale. By
the delta method, an estimate of this standard error is given by
\begin{equation*}
  \sigma^{\log}_n = \nabla g(\psi_n)^T \Sigma_n \nabla g(\psi_n),
\end{equation*}
where $\Sigma_n$ is the doubly-robust covariance matrix estimate output from
`drtmle` and $\nabla g(\psi)$ is the gradient of $\log\{\psi(1)/\psi(0)\}$. This
confidence interval may be obtained as follows.

```{r}
riskRatio <- list(f = function(eff){ log(eff) },
                  f_inv = function(eff){ exp(eff) },
                  h = function(est){ est[1] / est[2] },
                  fh_grad =  function(est){ c(1 / est[1], -1 / est[2]) })
ci(npreg_fit, contrast = riskRatio)
```

More generally, this method of inputting the `contrast` option to the `ci`
method can be used to compute confidence intervals of the form
\begin{equation*}
  f^{-1}\bigl\{ f(h(\psi_n)) \pm z_{1-\alpha/2} \nabla f(h(\psi_n))^T \Sigma_n
  \nabla f(h(\psi_n)) \bigr\} \ ,
\end{equation*}
where $f$ (`contrast$f`) is the transformation of the confidence interval,
$f^{-1}$ (`contrast$f_inv`) is the back-transformation of the interval, $h$
(`contrast$h`) is the contrast of the marginal means, and $\nabla f(h(\psi_n))$
(`contrast$fh_grad`) defines the gradient of the transformed contrast at the
estimated marginal means.

We note that this method of computing confidence intervals can also be used to
obtain a transformed confidence interval for a particular marginal mean. For
example, recall that $Y$ is binary in our running example. Thus, we may wish to
enforce that our confidence interval be computed on a scale ensuring that the
estimated confidence limits fall within the unit interval $[0, 1]$. To that end,
we can utilize an interval of the form:
\begin{equation*}
  \expit \biggl[ \log\biggl\{ \frac{\psi_n(1)}{1 - \psi_n(1)} \biggr\}
  \pm z_{1-\alpha/2} \sigma^{\text{logit}}_n \biggr] \ .
\end{equation*}
This confidence interval may be implemented as follows.

```{r}
logitMean <- list(f = function(eff){ qlogis(eff) },
                  f_inv = function(eff){ plogis(eff) },
                  h = function(est){ est[1] },
                  fh_grad = function(est){ c(1 / (est[1] - est[1]^2), 0) })
ci(npreg_fit, contrast = logitMean)
```

@@

As with the `ci` method, the `wald_test` method allows users to test linear
combinations of marginal means by inputting a vector of ones and negative ones
in the `contrast` option. We can use this to test the null hypothesis of no
average treatment effect or to test the null hypothesis that the average
treatment effect equals a given reference value, e.g., $0.05$, by specifying
the `null` option.

@@

The `wald_test()` method also allows for user-specified contrasts to be tested
using syntax similar to the `ci()` method. The only difference syntactically is
that it is not strictly necessary to specify `f_inv`, as the hypothesis test is
performed on the transformed scale. That is, we can generally test the null
hypothesis that $f(h(\psi_0))$ equals $f_0$ (the function $f$ applied to the
value passed to `null`) using the Wald statistic
\begin{equation*}
  Z_n \coloneqq \frac{\{f(h(\psi_n)) - f_0\}}{\nabla f(h(\psi_n))^T \Sigma_n
  \nabla f(h(\psi_n))} \ .
\end{equation*}
We can use the `riskRatio` contrast defined previously to test the null
hypothesis that $\psi_0(1) / \psi_0(0) = 1$.

@@

The `drtmle()` function supports missing data in `A` and `Y`. Such missing data
are common in practice and, if ignored, can bias estimates of the marginal
means of interest. The estimators previously discussed can be modified to allow
such missingness processes to depend on observed covariates. Let $\Delta_A$ and
$\Delta_Y$ be indicators that $A$ and $Y$ are observed, respectively. We can
redefine the OR to be $\bar{Q}_0(a,w) \coloneqq E_0(\Delta_Y Y \mid \Delta_A
A = a, \Delta_A = 1, \Delta_Y = 1, W = w)$ and similarly redefine the PS to be
\begin{align}
  g_0(a \mid w) &= Pr_0(\Delta_A = 1, \Delta_A A = a, \Delta_Y = 1 \mid W = w)
  \notag \\
  &= Pr_0(\Delta_A = 1 \mid W) Pr_0(\Delta_A A = a \mid \Delta_A = 1, W = w)
  \label{modPS} \\
  &\hspace{1.5in} Pr_0(\Delta_Y = 1 \mid \Delta_A = 1, \Delta_A A = a, W = w) \
   \notag.
\end{align}

To demonstrate, we artificially introduce missing values to `A` and `Y` in our
running example.

@@

The only modification of the call to `drtmle()` is in how the PS estimation is
performed. The options `glm_g` and `SL_g` are still used to fit generalized
linear models or a super learner, respectively. However, the code allows for
separate regression fits for each of the three components of the PS in (8). To
perform separate generalized linear model fits for each of the three components,
`glm_g` should be a named list with entries `"DeltaA"`, `"A"`, and `"DeltaY"`.
Respectively, these should provide a formula expression for the regression of
$\Delta_A$ on $W$, $A$ on $W$ in observations with $\Delta_A = 1$, and of
$\Delta_Y$ on $W$ and $A$ in observations with $\Delta_A = \Delta_Y = 1$ (if
`stratify = FALSE`) _or_ of $\Delta_Y$ on $W$ to be fit in observations with
$\Delta_A = \Delta_Y = 1$ and $A = a$ for each $a$ specified in option `a_0` (if
`stratify = TRUE`). If only a single formula is passed to `glm_g`, it is used
for each of the three regressions. We illustrate a call to `drtmle()` with
missing data and generalized linear model fits below.

@@

It is possible to mix generalized linear models and super learning when fitting
each of the components of the PS. In the example below, we use the wrapper
function `SL.glm`, which fits a main terms logistic regression, to fit the
regression of $\Delta_A$ on $W$; use `SL.npreg` to fit the regression of $A$ on
$W$; and use a super learner with library `SL.glm`, `SL.mean`, and `SL.earth` to
fit the regression of $\Delta_Y$ on $W$ and $A$.

@@

We can also fit the PS regressions externally and pass in via `gn`, though this
process is slightly more complicated than when no missing data are present. The
basic idea is to fit one regression for (i) the indicator of missing $A$, (ii)
$A$ itself, and (iii) the indicator of the outcome being missing. The three PS's
are then multiplied together and an appropriately ordered list is constructed as
before. We illustrate with the following simple example.

@@

In our examples so far, we have only considered binary treatments. However,
`drtmle` is equipped to handle treatments with an arbitrary number of discrete
values. Suppose that $A$ assumes values in a set $\mathcal{A}$, the PS
regression estimation is modified to ensure that
\begin{equation*}
  \sum\limits_{a \in \mathcal{A}} g_n(a \mid w) = 1 \ \mbox{for all $w$} \ .
\end{equation*}
If this condition is true, we say that the estimates of the PS are compatible.
Many regression methodologies that have been developed work well with binary
outcomes, but have not been extended to multi-level outcomes. To ensure that
users of `drtmle` have access to the large number of binary regression
techniques when estimating the PS, we have taken an alternative approach to
estimation of the propensity for each level of the treatment. Specifically,
rather than fitting a single multinomial-type regression, we fit a series of
binomial regressions. As a concrete example, consider the case of no missing
data and where $\mathcal{A} = \{0,1,2\}$. We can ensure compatible estimates of
the PS by generating an estimate $g_n(0 \mid w)$ of $Pr_0(A = 0 \mid W = w)$
and $\tilde{g}_n(1 \mid w)$ of $Pr_0(A = 1 \mid A > 0, W = w)$. The latter
estimate may be generated by regression $I(A = 1)$ on $W$ in observations with
$A > 0$. Because the outcome is binary, this regression may be estimated using
any technique suitable for binary outcomes. By simple rules of conditional
probability, we know that $g_0(1 \mid w) = Pr_0(A = 1 \mid A > 0, W = w)\{1 -
Pr_0(A = 0 \mid W)\}$ and thus an estimate of $g_0(1 \mid w)$ can be computed
as $g_n(1 \mid w) = \tilde{g}_n(1 \mid w) \{1 - g_n(0 \mid w)\}$. Finally, we
let $g_n(2 \mid w) = 1 - g_n(0 \mid w) - g_n(1 \mid w)$, which ensures
compatibility of the estimates for each level of covariates. This approach
generalizes to an arbitrary number of treatment levels. Below, we provide an
illustration using simulated data with a treatment that has three levels. The
true parameter values in this simulation are $\psi_0(0)=$ `r round(EY0,2)`,
$\psi_0(1)=$ `r round(EY1, 2)`, and $\psi_0(2)=$ `r round(EY2,2)`.
